Accurate Simulation of Flow through Dipping Aquifers with MODFLOW 6 Using Enhanced Cell Connectivity

### 2024-12-05 latex/article_clean_FINAL - Copy.tex
last-minute tweaks made before resubmission
last-minute tweaks made before resubmission

The solid blue vectors in Figure \ref{fig:schem-conn-area-flux}b show that to simulate steady, uniform flow along the dipping hydrogeologic layer, cells must exchange water not only ``horizontally'' with neighboring cells in the same grid layer, but also vertically with cells in the adjacent grid layer. Specifically, there must be flow from cells in the lower grid layer to cells in the upper grid layer, which corresponds to the vertical component of the groundwater flux. Note, however, that the solid blue vectors alone are incompatible with steady, uniform flow along the hydrogeologic layer because the exchange of water between the two grid layers is in one direction only: upward. Such a steady-state flow configuration would imply continuous depletion of flow within the lower grid layer and accumulation of flow within the upper grid layer as one moves along the hydrogeologic layer in the direction of flow. Thus, the steady, uniform flow represented by the gray vectors cannot be simulated accurately without some mechanism for returning flow from the upper grid layer to the lower grid layer. The full-connectivity grid provides such a mechanism by offering additional connections between cells in adjacent grid layers. Vertical flows from the lower grid layer to the upper grid layer, represented by vertical solid blue vectors in Figure \ref{fig:schem-conn-area-flux}b, can be returned to the lower grid layer by nominally ``horizontal'' flows, represented by dashed red vectors in Figure \ref{fig:schem-conn-area-flux}b. Although this concept has been illustrated using a two-grid-layer model for simplicity, the same reasoning applies given any number of grid layers.

@@

A link to the Jupyter notebooks and MODFLOW version 6.5.0 executable code used to create and run the test problem is available in the Supporting Information section. The notebooks leverage the capabilities of FloPy \citep{bakker2016scripting, hughes2024flopy} to assist in setting up input for and processing results from the MODFLOW 6 simulations. A Python script for converting structured, DIS grids to unstructured, DISU grids with full connectivity is included.

@@

\section*{Acknowledgments}
Kerry Bardot is supported by The Australian Research Council, The Western Australian Department of Water and Environmental Regulation and The Water Corporation of Western Australia through grant number LP18101153. The authors sincerely thank James Rumbaugh, John Masterson, Kalle Jahn, Jacob LaFontaine, John Brakebill, Heather Welch, Randall Hunt, and Richard Yager for their review comments, which improved the clarity of the manuscript and refined our conclusions regarding the practical implications of this study. Any use of trade, firm, or product names is for descriptive purposes only and does not imply endorsement by the U.S. Government.

\section*{Supporting Information}
The files that support the creation and running of the test problem are available at https://doi.org/10.5066/P13BNARA.

>>>>

The solid blue vectors in Figure \ref{fig:schem-conn-area-flux}b show that to simulate steady, uniform flow along the dipping hydrogeologic layer, cells must exchange water not only ``horizontally'' with neighboring cells in the same grid layer, but also vertically with cells in the adjacent grid layer. Specifically, there must be flow from cells in the lower grid layer to cells in the upper grid layer, which corresponds to the vertical component of the groundwater flux. Note, however, that the solid blue vectors alone are incompatible with steady, uniform flow along the hydrogeologic layer because the exchange of water between the two grid layers is in one direction only: upward. Such a steady-state flow configuration would imply continuous depletion of flow within the lower grid layer and accumulation of flow within the upper grid layer as one moves along the hydrogeologic layer in the direction of flow. Thus, the steady, uniform flow represented by the solid gray vectors cannot be simulated accurately without some mechanism for returning flow from the upper grid layer to the lower grid layer. The full-connectivity grid provides such a mechanism by offering additional connections between cells in adjacent grid layers. Vertical flows from the lower grid layer to the upper grid layer, represented by vertical solid blue vectors in Figure \ref{fig:schem-conn-area-flux}b, can be returned to the lower grid layer by nominally ``horizontal'' flows, represented by dashed red vectors in Figure \ref{fig:schem-conn-area-flux}b. Although this concept has been illustrated using a two-grid-layer model for simplicity, the same reasoning applies given any number of grid layers.

@@

A link to the Jupyter notebooks \citep{jupyterproject} and MODFLOW version 6.5.0 executable code used to create and run the test problem is available in the Supporting Information section. The notebooks leverage the capabilities of FloPy \citep{bakker2016scripting, hughes2024flopy} to assist in setting up input for and processing results from the MODFLOW 6 simulations. A Python \citep{pythonsoftware} script for converting structured, DIS grids to unstructured, DISU grids with full connectivity is included.

@@

\section*{Acknowledgments}
Kerry Bardot is supported by The Australian Research Council, The Western Australian Department of Water and Environmental Regulation and The Water Corporation of Western Australia through grant number LP18101153. The authors sincerely thank James Rumbaugh, John Masterson, Kalle Jahn, Jacob LaFontaine, Randall Hunt, and Richard Yager for their review comments, which improved the clarity of the manuscript and refined our conclusions regarding the practical implications of this study. Any use of trade, firm, or product names is for descriptive purposes only and does not imply endorsement by the U.S. Government.

\section*{Supporting Information}
The files that support the creation and running of the test problem are available at https://doi.org/10.5066/P13BNARA \citep{suppinfodatarelease}.
